Combination of field and experimental data together with computational models

### 2022-02-16 manuscript_1.0/manuscript_1.0.Rmd
Fixed data files. Updated the cpp to g++ compilation. new parameter files to run chains with the fixed data files. Update figures in the manuscript

### Overview

In this study, we combined empirical data from Triki *et. al* 
[@triki_Biological_2019;@triki_Brain_2020] and the computational 
model analysed in Quiñones *et al.*[@quinones_Reinforcement_2019] 
to test the relative validity of chaining versus operant conditioning as 
mechanisms to account for future consequences of cleaner fish desicion-making.
In order to compare the explanatory power of the different hypotheses, 
we estimated the parameter values of three versions of 
the model that best fit the empirical data. 
Particularly, we fitted the parameter that determines how important 
future reward is ($\gamma$), the magnitude of the negative reward 
($\eta$) and a scaling constant that translates cleaner fish absolute 
to relative population abundances (see following sections for details). 
We obtained the model predictions using a set of stochastic individual-based
simulations. Thus, each parameter combination of the model yielded not one 
but a distribution of predictions. For the 
fitting procedure, we used the bayesian inference algorithm: 
Markov Chain Monte Carlo (MCMC). MCMC is a stochastic algorithm that allows
to sample from the posterior distribution of model parameter values. The posterior
distribution represents the most likely parameter values given the data. 
Due to the stochastic nature of MCMC, which repeatedly samples parameter values
and evaluates their likelihood, the distribution of likelihoods from the 
stochastic model is also explored. This makes MCMC and bayesian inference
particularly suitable for the computational models we evaluate. Bellow, 
we explain first the computational model and the data collection, and later 
how the MCMC was used to estimate parameter values.

@@

In order to compare the fit that the three alternative models give we use the 
distribution of pseudo $R^2$ proposed by Mcfadden [@mcfadden_Conditional_1974].
Mcfadden's pseudo $R^2$ is a standard measure of fit for logistic regression. 
In that context, $pseudo-R^2$ uses the log-likelihood of the data given 
the model, relative to the log-likelihood of the data given a 
model without covariates, as a measure of fit. Given that our 
model is not a logistic regression, we measure the $pseudo-R^2$ in
relation to the log-likelihood of a model
with parameters $\gamma$ and $\eta$ set to zero. This in practice
amounts to a model that has neutral preference between the two options. 
Pseudo $R^2$ was calculated for all the samples of the posterior gotten from
the MCMC. Thus, we use these distributions of pseudo $R^2$'s as a measure of
fit.

@@

```{r post, fig.cap="Posterior distributions for parameter values $\\gamma$, $\\eta$ and the scaling constant. We show the kernel density estimates, below the mode (black point) and the 65\\% (light blue) and 95\\% (grey)  highest posterior density interval for the three paramteres. On the left (panels a, c and c) posterior distributions for a model with all three parameters; while on the right (b and d) a model without negative reward. Panel f shows a measure of fit for both models, namely the distribution of pseudo-$R^2$ obtained from sampling the posterior distribution of parameter values.",out.width='100%',fig.height=6}
library(data.table)
library(coda)
library(ggplot2)
library(cowplot)
library(ggdist)
source(here("../R_files/posPlots.R"))
source(here("loadData.R"))
library("jsonlite")

@@

# NLables names for the three scenarios
labels.Scen<-c("both","gam","Nrew")
labelsPlot.Scen<-c("Full \n model","Future \n reward","Negative \n reward")

@@

panel.rsqr.all<-
  ggplot(data=loglikehoods.all[is.finite(lglikelihood)],
                  aes(y=model,x=Rsqrd,fill=model))+
  stat_halfeye(alpha=0.5,point_size=3)+
  theme_classic()+xlim(-0.6,0.2)+
  scale_fill_manual(values=c("#1b9e77","#d95f02","#7570b3"),
  name = "Model", labels = labelsPlot.Scen)+
  geom_vline(xintercept = 0,color="black",size=1)+
  theme(legend.position = c(.3, .75),legend.key.size = unit(0.25,'cm'),
        axis.text.y = element_blank())

plot_grid(nrow=3,align = "v",byrow = FALSE,
          # Full model - \gamma
          ggplot(data=MCMCdata1,aes(x=gamma)) + 
            stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.1$gamma))),
                         point_interval = mode_hdi, .width = c(.66, .95),point_size=3,
                         show.legend=FALSE) + 
            labs(title=labelsPlot.Scen[1],subtitle = expression(gamma),x="",y="")+
            theme_classic()+scale_fill_manual(values=c("gray85","skyblue","gray85"))+
            theme(plot.margin = unit(rep(0.5,4), "lines")),
          # \eta
          ggplot(data=MCMCdata1,aes(x=negReward)) + 
            stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.1$negReward))),
                         point_interval = mode_hdi, .width = c(.66, .95),point_size=3,
                         show.legend=FALSE)+theme_classic()+
            scale_fill_manual(values=c("gray85","skyblue","gray85"))+
            labs(subtitle = expression(eta),x="",y="")+xlim(-5,5)+
            theme(plot.margin = unit(rep(0.5,4), "lines")),
          # Scale Constant
          ggplot(data=MCMCdata1,aes(x=scaleConst)) + 
            stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.1$scaleConst))),
                         point_interval = mode_hdi, .width = c(.66, .95),point_size=3,
                         show.legend=FALSE)+
            theme_classic()+scale_fill_manual(values=c("skyblue","gray85"))+
          labs(subtitle = "Scalling const.",x="",y="")+xlim(-2,1500)+theme_classic()+
          theme(plot.margin = unit(rep(0.5,4), "lines")),
          # Future reward only model - \gamma
          ggplot(data=MCMCdata2,aes(x=gamma)) +
            stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.2$gamma))),
                         point_interval = mode_hdi, .width = c(.66, .95),point_size=3,
                         show.legend=FALSE) +
            scale_fill_manual(values=c("gray85","skyblue","gray85"))+
            labs(title=labelsPlot.Scen[2],subtitle = expression(gamma),x="",y="")+
            theme_classic()+theme(plot.margin = unit(rep(0.5,4), "lines")),
          # Scalling constant
          ggplot(data=MCMCdata2,aes(x=scaleConst)) +
            stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.2$scaleConst))),
                         point_interval = mode_hdi, .width = c(.66, .95),point_size=3,
                         show.legend=FALSE) +
            scale_fill_manual(values=c("gray85","skyblue","gray85"))+
            labs(subtitle = "Scalling const.",x="",y="")+
            theme_classic()+theme(plot.margin = unit(rep(0.5,4), "lines")),
          # Pseudo R^2
          panel.rsqr.all+labs(subtitle = expression(pseudo-R^2),x="",y=""),
          labels=c('a','b','c','d','e')
)

@@

Figure \ref{fig:post} panels a,c and e show the posterior distributions
of the parameter values of the full model, which includes both future
and negative reward. Posterior distributions show the probability
density (y axis) of the estimated parameter values (x axis). Notably,
the bulk of the posterior distribution for the parameter tuning negative
reward ($\eta$ panel e) is around zero. Thus, the shape of the posterior
and its credible intervals suggest the absence of negative reward to be
well supported by the data. As for $\gamma$, the confidence intervals
also includes zero, but the mode of the posterior is around 0.5. Thus,
we run the analysis setting $\eta$ to zero. Figure \ref{fig:post} panels
b and d, show the posterior distributions of the two parameters
estimated. Not surprisingly, under a model without negative reward, the
distribution of $\gamma$ shifts to higher values. Under the new model,
the estimates of $\gamma$ do not include zero and the distribution is
centered around $0.5$. In panel f of figure \ref{fig:post} 
we show the distribution of $pseudo-R^2$ calculated using 
samples from the posterior distributions shown before, 
as well as from a model with negative reward but without
future reward. Note, $pseudo-R^2$ can have negative values, that is when
the log-likelihood of the model is lower than that of a model that
triggers neutral preferences. Even though, the peak of the three
$pseudo-R^2$ distributions is not very different, the model with only
future reward produces a distribution of $pseudo-R^2$ where more values
are positive (to the right of black line in Fig. \ref{fig:post} f). This
shows that, accounting for variation in the parameter estimates, the
model with only future reward gives a better fit to the data, despite
having one parameter less.

```{r pred, fig.cap="Observed and predicted probability of choosing a visitor. Left-hand side panel: color contour shows the prediction of the learning model using the mode of the posterior distributions of parameters recovered by the statistical analysis. Dots show the frequency of visitor choices for the 12 locations, as well as the corresponding relative cleaner abundance (x axis) and frequency of visitors leaving the station (y axis). Right-hand side panels: Variation of the predicted probabilites  of choosing a visitor and their observed values for 12 locations. Circles show the mean prediction for each location from 100 samples taking from the posterior distribution. Thick and thin bars show the 66 and 95\\% credible interval taken from those posterior samples. Squares show the predictions used for panel on the righ-hand side. Colour denotes different locations (see Empirical data section). Black line correspond to a perfect match between observed and predicted probabilities. Upper panels show prediction from a model including future and negative reward, lower panels from a  model with only future reward.",out.width='100%',fig.height=5}

@@

Figure \ref{fig:pred} shows predictions from the two models. On the
left, the colour coded contour shows the prediction of the model for
combinations of cleaner abundance (x axis) and visitor leaving
probability (y axis). Predictions were generated by running the model
using the mode of the posterior distributions as the parameter values,
and varying systematically cleaner abundance and visitor leaving
probability. Colour of the points on top of the contour show the average
frequency with which cleaner fish chose the visitor plate in their last
20 experimental trials. Each point corresponds to a location with a
combination of cleaner abundance and visitor leaving probability. Panels
on the right show how close the observed values for the probability of
choosing the visitor (y axis) are from the predictions of the models (x
axis). The black line represents the perfect fit between observed and
predicted values. Bars show the variation in predicted values using 100 
samples from the posterior distribution. Despite having one more 
parameter varying, the full model
gives a lower fit to the data than the model without negative reward.
In the supplementary material, we also show predictions from 
model with only negative reward, excluding future rewards ($\gamma=0$,
Fig. \ref{fig:nogamma}), provides a lower fit than either of the ones
presented in the main text.

The main reason for negative and future reward to give different
predictions is the way that cleaner's relative abundance influences the
preference for the visitor clients. Visitor leaving probability has a
similar positive effect on the probability of choosing the visitor
clients on all three models (Fig. \ref{fig:pred}, \ref{fig:nogamma}). In
contrast, cleaner's relative abundance has a different effect in the
model with only future reward, compared to the other two. In the model
with only future reward only intermediate cleaner abundance triggers a
preference for the visitor clients (Fig. \ref{fig:pred} c). In models
with negative reward, both intermediate and low cleaner abundance
trigger a preference for the visitor. (Fig. \ref{fig:pred} a,
\ref{fig:nogamma}).

@@

We asked which if two potential cognitive mechanisms 
– chaining of events, negative reinforcement, or a combination of both – 
is the more likely mechanism to explain intra-specific variation in cleaner 
performance in the ephemeral reward task as a function of ecological 
conditions in the wild. To evaluate the merits of the alternative mechanisms, 
we considered performance in the task to depend on decision-rules that 
cleaners had developed in their natural environment: individuals that 
solved the task did so because they already had a general preference for 
visitors, and applied this rule once being familiar with the task.  
While all three models capture well the positive relation between visitor 
leaving behaviour and cleaner performance in
the market task [@triki_Biological_2019], only the chaining mechanism 
predicts that cleaner performance in the task should be low if their 
local abundance in the wild relative to clients is low. Indeed, such low 
performance is predicted regardless of the visitor leaving
probability. In contrast, a model with negative reward predicts the
highest performance in the ephemeral reward task when relative 
cleaner abundance is low, particularly
together with high probability of visitor leaving (Fig. \ref{fig:pred}).
Low cleaner abundances mean the market has an excess of demand for
cleaning services. In the model, this translates to a cleaning station
that is frequently full. Thus, when visitors leave, it's likely that the
cleaner will have access to another client in the next step. Therefore, 
there will not be much difference in future reward between choosing 
a visitor and a resident, and cleaners will not develop a preference for the
visitor in these conditions. On the other hand, the effect of negative reward 
would be opposite as in a full station, cleaners will get more often the
heterotypic (resident-visitor) option and will develop a preference for
the visitor faster. Note that when cleaner abundance is high situations
with a resident and a visitor will be so rare that neither mechanism
will be very efficient in generating a preference for visitors. When
such choice arises it is still be best to choose the visitor; however,
the learning machinery will not be able to develop this preference
efficiently. Overall, estimation of future reward is then the cognitive
mechanism that allows cleaners to adaptively respond to the
ecological conditions of the biological market.

Our results support chaining as the more plausible mechanism behind the 
inter-specific variation in cleaners' performance, which is also 
the mechanism that allows them to respond more adaptively. This could
suggest chaining to be under positive selection. In the standard ephemeral
reward task, however, food amount and quality are equal between the 
two plates. An given that models resemble this, choosing visitors over 
residents in low cleaner abundances does not diminish the 
amount of food/reward in the the models. 
On the other hand, a cleaner committed to visitors might find more difficult 
to include in the decision making other variables, like parasite load 
and/or mucus quality, that do predict the amount of food in the field 
[@roche_Client_2021]. In nature and in laboratory experiments, cleaners prefer 
higher quality mucus and more food in the absence of market effects 
(i.e. in simultaneous 2-choice options, 
[@triki_Fluctuations_2019;@grutter_Cleaner_2004]), and they use size as a 
proxy for food availability [@wismer_Cuebased_2019;@grutter_Does_2005]. 
How important this limitation is will determine whether chaining yields a higher
fitness for cleaners.

Individual cleaners that performed in the laboratory test according to what 
is best at their site of capture (low performance when living at low densities, 
high performance when living at high densities) had larger forebrains 
harbouring more cells than individuals whose performance mismatched locally 
optimal decision-rules [@triki_Brain_2020]. Triki *et al.*  refer to the 
former as socially competent cleaners. Social competence refers to the 
ability to optimise social behaviour depending on the available social 
information [@taborsky_Social_2012a;
@bshary_Cooperation_2015;@varela_Correlated_2020]. Our analyses yielded no 
evidence that the difference in social competence with respect to the local 
market conditions and associated brain morphology are due to the mechanism 
used to incorporate future consequences. It could have been conceivable that 
high performing individuals from low-density sites might use negative 
reinforcement instead of chaining, but in that case negative reinforcement 
should have explained part of the data. Configurational learning or 
chunking [@sutherland_Configural_1989;@miller_Magical_1956], the second component 
necessary to solve the ephemeral reward task [@quinones_Reinforcement_2019], was 
not varied in the models analized here. However, while chunking tendencies 
should vary to allow individuals to adapt to local conditions 
[@prat_Modelling_2022;@kolodny_Evolution_2014], systematic differences 
in individual chunking tendencies would not explain how socially 
competent decisions vary as a function of relative abundance. 
Therefore, it remains currently unclear what cell-demanding 
mechanisms may cause variation in social competence that 
translates into site-specific variation in performance in the 
ephemeral reward task.

The model is inspired by the general processes of associative learning
where short term rewards are translated to decision making; thus, it 
ignores alternative channels of information that could be relevant in 
market-like situations. For example, the model does not investigate whether 
cleaners actually assess the frequency of client visits or a mean frequency 
of visitors leaving. The updating learning mechanism for the development 
of preferences works on a trial by trial basis. In the model, cleaners 
do not need to assess the actual state of the market, *i.e.* their density, 
the density of residents and visitors, and client visitation rate as an 
indicator of demand. They only need to assess short-term consequences of 
own decisions on food intake and chain them. Also, for the sake of 
simplicity, the model ignores the process by which cleaners discriminate 
residents and visitor clients. A model that accounts for this discrimination
probably involves the development of preferences for morphological or 
behavioural features that are statistically associated with visitors 
or residents. For example, visitors are on average larger larger 
than residents [@bshary_Cleaner_2001a], and are less likely to chase 
cleaners in response to the latter cheating by taking a bite of 
mucus [@bshary_Asymmetric_2002]. Given these associations, chaining might 
produce the decision-rule “chose the larger client and/or less aggressive 
client”, which does not help in the standard ephemeral reward task.

In conclusion, our study shows that variation in cognitive performance as 
a function of ecological factors may set the stage for the use of 
mechanistic modelling to identify the cognitive processes underlying 
learning in animals. The combination overcomes the limitations 
of the general philosophy in animal cognition to apply the logic of 
Lloyd Morgen's canon (Occam's razor). Excluding basic 
reinforcement learning explanations (operant and/or classical conditioning) 
as a potential explanation in experiments
often warrants animals solving the task on the first possible occasion.
For example, any theory of mind task needs to be solved in the first
trial in order to exclude fast conditioning [@heyes_Theory_1998].
Similarly, subjects need to solve a social learning task on the first
trial to accept imitation as a mechanism over stimulus/local
enhancement. Such strict conditions are virtually never met. For
example, potato washing by Japanese macaques, an iconic example of
social learning, took several years to spread within the group
[@kawamura_Process_1959], meaning that any learner had been repeatedly
exposed to demonstrations before acquisition. Importantly, Galef
[@galef_Question_1992] refuted imitation as mechanism not simply because
of the repeated exposure but because a (rather qualitative) analysis of
the spread of potato washing across individuals did not follow the
prediction based on imitation learning (see also
[@hirata_SweetPotato_2001]). In our case, the number of trials it took
cleaners to learn the solution to the market task would never allow to
exclude an important role of negative reinforcement based on the data
alone. However, fitting model predictions to our data set revealed that
a more complex mechanism (estimation of future reward) fits the data
better.

>>>>

<!-- ### Overview -->

<!-- In this study, we combined empirical data from Triki *et. al*  -->
<!-- [@triki_Biological_2019;@triki_Brain_2020] and the computational  -->
<!-- model analysed in Quiñones *et al.*[@quinones_Reinforcement_2019]  -->
<!-- to test the relative validity of chaining versus operant conditioning as  -->
<!-- mechanisms to account for future consequences of cleaner fish desicion-making. -->
<!-- In order to compare the explanatory power of the different hypotheses,  -->
<!-- we estimated the parameter values of three versions of  -->
<!-- the model that best fit the empirical data.  -->
<!-- Particularly, we fitted the parameter that determines how important  -->
<!-- future reward is ($\gamma$), the magnitude of the negative reward  -->
<!-- ($\eta$) and a scaling constant that translates cleaner fish absolute  -->
<!-- to relative population abundances (see following sections for details).  -->
<!-- We obtained the model predictions using a set of stochastic individual-based -->
<!-- simulations. Thus, each parameter combination of the model yielded not one  -->
<!-- but a distribution of predictions. For the  -->
<!-- fitting procedure, we used the bayesian inference algorithm:  -->
<!-- Markov Chain Monte Carlo (MCMC). MCMC is a stochastic algorithm that allows -->
<!-- to sample from the posterior distribution of model parameter values. The posterior -->
<!-- distribution represents the most likely parameter values given the data.  -->
<!-- Due to the stochastic nature of MCMC, which repeatedly samples parameter values -->
<!-- and evaluates their likelihood, the distribution of likelihoods from the  -->
<!-- stochastic model is also explored. This makes MCMC and bayesian inference -->
<!-- particularly suitable for the computational models we evaluate. Bellow,  -->
<!-- we explain first the computational model and the data collection, and later  -->
<!-- how the MCMC was used to estimate parameter values. -->

@@

To compare the fit of the three alternative models, we used the 
distribution of pseudo $R^2$ proposed by Mcfadden [@mcfadden_Conditional_1974].
Mcfadden's pseudo $R^2$ is a standard measure of fit for logistic regression. 
In that context, $pseudo-R^2$ uses the log-likelihood of the data given 
the model, relative to the log-likelihood of the data given a 
model without covariates, as a measure of fit. Our 
model is not a logistic regression, therefore we measured the $pseudo-R^2$ in
relation to the log-likelihood of a model
with parameters $\gamma$ and $\eta$ set to zero. This in practice
amounts to a model that has neutral preference between the two options. 
We computed the pseudo $R^2$ for all the samples of the posterior from
the MCMC. Thus, we used these distributions of pseudo $R^2$'s as a measure of
fit.

@@

```{r post, fig.cap="Posterior distributions for parameter values $\\gamma$, $\\eta$ and the scaling constant. We show the kernel density estimates, below the mode (black dot) and the 65\\% (light blue shade) and 95\\% (grey shade)  highest posterior density interval for the three paramteres. On the left, panels a, b and c show posterior distributions for a model with all three parameters; while on the right, d and e, show a model with only chaining. Panel f shows a measure of fit for both models, namely the distribution of pseudo-$R^2$ obtained from sampling the posterior distribution of parameter values.",out.width='100%',fig.height=6}
library(data.table)
library(coda)
library(ggplot2)
library(cowplot)
library(ggdist)
source(here("../R_files/posPlots.R"))
source(here("loadData.R"))
library("jsonlite")

@@

# NLables names for the three scenarios
labels.Scen<-c("both","gam","Nrew")
labelsPlot.Scen<-c("FM","C","OC")

@@

panel.rsqr.all<-
  ggplot(data=loglikehoods.all[is.finite(lglikelihood)],
                  aes(y=model,x=Rsqrd,fill=model))+
  stat_halfeye(alpha=0.5,point_size=3)+
  theme_classic()+xlim(-0.5,0.25)+
  scale_fill_manual(values=c("#1b9e77","#d95f02","#7570b3"),
  name = "Model", labels = labelsPlot.Scen)+
  scale_y_discrete(labels=labelsPlot.Scen)+
  geom_vline(xintercept = 0,color="black",size=0.5)+
  theme(legend.position = 'none')
          # c(.32, .65)
        # ,legend.key.size = unit(0.22,'cm'),
        # axis.text.y = element_blank(),
        # legend.text = element_text(size=7))

gam.both.post<-ggplot(data=MCMCdata1,aes(x=gamma)) + 
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.1$gamma))),
               point_interval = mode_hdi, .width = c(.66, .95),
               point_size=3,show.legend=FALSE) + 
  labs(title=labelsPlot.Scen[1],
       subtitle = expression(gamma),x="",y="")+
  theme_classic()+
  scale_fill_manual(values=c("gray85","skyblue","gray85"))
nrew.both.post<-ggplot(data=MCMCdata1,aes(x=negReward)) + 
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.1$negReward))),
               point_interval = mode_hdi, 
               .width = c(.66, .95),point_size=3,
               show.legend=FALSE)+theme_classic()+
  scale_fill_manual(values=c("gray85","skyblue","gray85"))+
  labs(title="",subtitle = expression(eta),x="",y="")+
  xlim(-2,2)
gam.gam.post<-ggplot(data=MCMCdata2,aes(x=gamma)) +
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.2$gamma))),
               point_interval = mode_hdi, .width = c(.66, .95),
               point_size=3,
               show.legend=FALSE) +
  scale_fill_manual(values=c("gray85","skyblue","gray85"))+
  labs(title=labelsPlot.Scen[2],
       subtitle = expression(gamma),x="",y="")+
  theme_classic()
nrew.nrew.post<-ggplot(data=MCMCdata3,aes(x=negReward)) +
  stat_halfeye(aes(fill=stat(cut(x,breaks = cuts.3$negReward))),
               point_interval = mode_hdi, .width = c(.66, .95),
               point_size=3,
               show.legend=FALSE) +
  scale_fill_manual(values=c("gray85","skyblue","gray85"))+
  labs(title=labelsPlot.Scen[3],
       subtitle = expression(gamma),x="",y="")+
  theme_classic()+xlim(-2,2)

first.col<-plot_grid(nrow=3,align = "v",byrow = TRUE,
          gam.both.post,gam.gam.post,nrew.nrew.post,           
          labels=c('a','c','d'))
second.col<-plot_grid(nrow = 2,align = "v",byrow = TRUE,
          nrew.both.post,rsqr.both,rel_heights = c(1,2),
          labels = c('b','e'))
plot_grid(ncol = 2,first.col,second.col)

@@

Figure \ref{fig:post} panels a,b and c show the posterior distributions
of the parameter values of the full model, which includes both chaining
and operant conditioning (negative reward). Posterior distributions show the probability
density (y axis) of the estimated parameter values (x axis). Notably,
the bulk of the posterior distribution for the parameter tuning negative
reward ($\eta$ panel b) was around zero. Thus, the shape of the posterior
and its credible intervals suggest the absence of negative reward to be
well supported by the data. As for $\gamma$, the confidence intervals
also includes zero, but the mode of the posterior is around 0.5. Thus,
we ran the analysis setting $\eta$ to zero. Figure \ref{fig:post} panels
d and e, show the posterior distributions of the two parameters
estimated. Not surprisingly, under a model without negative reward, the
distribution of $\gamma$ shifted to higher values. Under the new model,
the estimates of $\gamma$ did not include zero and the distribution was
centred around $0.5$. In panel f of figure \ref{fig:post} 
we show the distribution of $pseudo-R^2$ calculated using 
samples from the posterior distributions shown before, 
as well as from a model with negative reward but without
future reward. Note, $pseudo-R^2$ can have negative values, which is when
the log-likelihood of the model is lower than that of a model that
triggers neutral preferences. Even though, the peak of the three
$pseudo-R^2$ distributions was not very different, the model with only
future reward produced a distribution of $pseudo-R^2$ where more values
were positive (to the right of black line in Fig. \ref{fig:post} f). This
shows that, accounting for variation in the parameter estimates, the
model with only future reward gives a better fit to the data, despite
having one parameter less.

```{r pred, fig.cap="Observed and predicted probability of choosing a visitor. Left-hand side panel: color contour shows the prediction of the learning model using the mode of the posterior distributions of parameters recovered by the statistical analysis. Dots show the frequency of visitor choices for the 12 reef sites, as well as the corresponding relative cleaner fish abundance (x axis) and frequency of visitors clients leaving the cleaning station (y axis). Right-hand side panels: Variation of the predicted probabilites  of choosing a visitor over a resident and their observed values for 12 locations. Circles show the mean prediction for each location from 100 samples taken from the posterior distribution. Thick and thin bars show the 66 and 95\\% credible interval, respectively, taken from those posterior samples. Squares show the predictions used for panel on the righ-hand side. Colour coding denotes different reef sites/year of the data collection (see Empirical data section). Black line corresponds to a perfect match between observed and predicted probabilities. Upper panels show prediction from a model including chaining and operant conditioning, lower panels from a  model with only chaining.",out.width='100%',fig.height=5}

@@

Figure \ref{fig:pred} shows predictions from the two models, 
one including chaining and operant conditioning (upper panels) and 
one with only chaining (lower panels). On the
left, the colour coded contour shows the prediction of the model for
combinations of cleaner fish abundance (x axis) and visitor leaving
probability (y axis). Predictions were generated by running the model
using the mode of the posterior distributions as the parameter values,
and varying systematically cleaner fish abundance and visitor leaving
probability. Colour of the dots on top of the contour show the average
frequency with which cleaner fish chose the visitor plate in their last
20 experimental trials. Each dot corresponds to a reef-site/year of 
data collection with a combination of cleaner fish abundance 
and visitor leaving probability. Panels
on the right show how close the observed values for the probability of
choosing the visitor (y axis) were from the predictions of the models (x
axis). The black line represents the perfect fit between observed and
predicted values. Bars show the variation in predicted values using 100 
samples from the posterior distribution. Despite having one more 
parameter varying, the full model
gave a lower fit to the data than the model without negative reward.
In the supplementary material, we also show predictions from 
model with only negative reward, excluding future rewards ($\gamma=0$,
Fig. \ref{fig:nogamma}), provides a lower fit than either of the ones
presented in the main text.

The main reason for chaining and operant conditioning to give different
predictions is the way that cleaner fish relative abundance influences the
preference for the visitor clients. Visitor leaving probability has a
similar positive effect on the probability of choosing the visitor
clients on all three models (Fig. \ref{fig:pred}, \ref{fig:nogamma}). In
contrast, cleaner fish relative abundance has a different effect in the
model with only chaining, compared to the other two models, 
full model and only operant conditioning. In the model
with only chaining, only intermediate cleaner fish abundance triggers a
preference for the visitor clients (Fig. \ref{fig:pred} c). In models
with negative reward, both intermediate and low cleaner abundance
triggered a preference for the visitor. (Fig. \ref{fig:pred} a,
\ref{fig:nogamma}).

@@

In this study, our main aim was to unravel which of two potential 
cognitive mechanisms, chaining of events, operant conditioning, 
or their combination, best explains wild-caught cleaner fish 
performance in the ephemeral reward task, while accounting for their
ecological conditions. To evaluate the merits of each of these two mechanisms
separately and combined, we considered cleaner fish performance 
in the lab test to have its origin from the these fish apply 
in their natural environment. That is, individuals that 
solved the task already had a preference for 
visitors clients, and generalized this rule to the lab conditions 
once being familiar with the task.

While all three models captured well the positive relation between visitor 
leaving behaviour and cleaner performance in
the market task [@triki_Biological_2019], only the chaining mechanism 
predicted that cleaner fish performance in the task should be low in 
habitats with low cleaner-to-client ratios, regardless of the 
visitor leaving probability. In contrast, models including 
negative reward predicted the highest performance in the ephemeral 
reward task when relative cleaner abundance is low, particularly 
together with high probability of visitor leaving (Fig. \ref{fig:pred}).
Low relative cleaner fish abundances mean the market has an excess of demand for
cleaning services. In the models, this translates to a cleaning station
that is frequently full. Thus, when a visitors leaves, it is likely that the
cleaner fish will have access to another client in the next step. Therefore, 
there will not be much difference in future reward between choosing 
a visitor and a resident, and cleaners will not develop a preference for the
visitor in these conditions. On the other hand, the effect of negative reward 
on cleaner fish preference is the opposite, as in a busy cleaning station, to
that of chaining. cleaner fish will get more often the resident-visitor
state and will develop a preference for the visitor faster. 
At high cleaner fish abundances, the resident-visitor state
becomes so rare that neither mechanism is very efficient 
at generating a preference for visitors. When facing the resident 
visitor choice it is still be best to choose the visitor; however,
the learning machinery will not be able to develop this preference
efficiently. Overall, the models suggest that chaining is the cognitive
mechanism that allows cleaner fish to adaptively adjust to their 
biological market ecological conditions.

Our results support chaining as the more plausible mechanism behind the 
inter-specific variation in cleaners' performance. This suggest
that chaining could be under positive selection. In the standard ephemeral
reward task, however, food amount and quality are equal in the 
two Plexiglas plates. Since our computational models simulate this aspect of
the task, choosing visitors over residents at low cleaner fish abundances 
does not diminish the amount of food/reward in the models. 
On the other hand, a cleaner fish committed to visitors might find it more difficult 
to include in its decision making other variables, like parasite load 
and/or mucus quality [@roche_Client_2021]. Indeed, in the absence of 
market effects, both field observations and aboratory experiments also 
using simultaneous two choice options, show that cleaner fish base their 
decision-making on other cues such as the mucus quality and the amount of food.  
[@triki_Fluctuations_2019;@grutter_Cleaner_2004], as well as client size 
as a proxy for food quantity [@wismer_Cuebased_2019;@grutter_Does_2005]. 
How important this limitation is will determine whether chaining yields a higher
fitness for cleaners.

Previous research showed that cleaner fish living at high population 
densities and giving service priority to the visitor plate in the 
ephemeral reward task, as well as cleaner fish living at low 
densities but denying service priority to the visitor plate possess 
larger forebrains; a key teleost brain region associated with 
behavioral flexibility and social intelligence. Those failing to 
show optimized decision-rules given their local ecological 
conditions had relatively smaller forebrains  [@triki_Brain_2020]. 
Triki *et al.*  refer to the former as socially competent cleaner fish,
while the second group as socially incompetent cleaner fish. 
Social competence is the ability to optimise social behaviour 
to the available social information [@taborsky_Social_2012a;
@bshary_Cooperation_2015;@varela_Correlated_2020]. Our analyses yielded no 
evidence that the difference in social competence with respect to the local 
ecological conditions and associated brain morphology, 
found by Triki *et al* [@triki_Brain_2020], are due to the mechanism 
used to incorporate future consequences. It could have been conceivable that 
high performing individuals from low population densities reef sites 
might use negative reinforcement instead of chaining, but in that 
case negative reinforcement should have explained at least 
part of the data. Configurational learning or 
chunking [@sutherland_Configural_1989;@miller_Magical_1956], 
the second component necessary to solve the ephemeral 
reward task [@quinones_Reinforcement_2019], was 
not varied in the models we analized here. However, while chunking tendencies 
should vary to allow individuals to adapt to their local conditions 
[@prat_Modelling_2022;@kolodny_Evolution_2014], systematic differences 
in individual chunking tendencies would not explain how socially 
competent decisions vary as a function of relative abundance. 
Therefore, it remains currently unclear what cell-demanding 
mechanisms may cause variation in social competence that 
translates into site-specific variation in performance in the 
ephemeral reward task.

Our models are inspired by the general processes of associative learning
where short term rewards are translated into decision making; thus, it 
ignores alternative channels of information that could be relevant in 
market-like situations. For example, the model does not investigate whether 
cleaner fish actually assess the frequency of client visits or a mean frequency 
of visitors leaving. The updating learning mechanism for the development 
of preferences works on a trial-by-trial basis. In the model, cleaner fish 
do not need to assess the actual state of the market, *i.e.* their abundance, 
the abundance of residents and visitors, and client visitation rate as an 
indicator of demand. They only need to assess short-term consequences of their
own decisions on food intake, and chain them. Also, for the sake of 
simplicity, the model ignores the process by which cleaner fish discriminate 
residents and visitor clients. A model that accounts for this discrimination
probably would involve the development of preferences for morphological or 
behavioural features that are statistically associated with visitors 
or residents. For example, visitors are on average larger 
than residents in body size [@bshary_Cleaner_2001a], and contrary to 
resindents, are less likely to chase a cleaner fish that fails to cooperate
and intead cheats its client by taking a bite of mucus 
[@bshary_Asymmetric_2002]. Given these associations, chaining might 
produce the decision-rule “choose the larger client and/or the less aggressive 
client”, which is not a useful rule in the standard ephemeral reward task.

In conclusion, our study shows that variation in cognitive performance as 
a function of the local ecological conditions may set the stage for the use of 
mechanistic modelling to identify the cognitive processes underlying 
learning in animals. The combination overcomes the limitations 
of the general philosophy in animal cognition to apply the logic of 
Lloyd Morgen's canon (Occam's razor). Cognitive experiments with the 
aim of excluding basic reinforcement learning as a potential 
explanation (operant and/or classical conditioning) of performance 
often employ one trial experiments requiring animals to solve the 
task on the first possible occasion. 
For example, any theory of mind task needs to be solved in the first
trial in order to exclude fast conditioning [@heyes_Theory_1998].
Similarly, subjects need to solve a social learning task on the first
trial to accept imitation as a mechanism over stimulus/local
enhancement. Such strict conditions are virtually never met. For
example, potato washing by Japanese macaques, an iconic example of
social learning, took several years to spread within the group
[@kawamura_Process_1959], meaning that any learner had been repeatedly
exposed to demonstrations before acquisition. Importantly, Galef
[@galef_Question_1992] refuted imitation as mechanism not simply because
of the repeated exposure but because a (rather qualitative) analysis of
the spread of potato washing across individuals did not follow the
prediction based on imitation learning (see also
[@hirata_SweetPotato_2001]). In our case, the number of trials it took
cleaners to learn the solution to the ephemeral rewards task 
would never allow to exclude an important role of 
negative reinforcement based on the data
alone. However, fitting model predictions to our comprehensive empirical 
data set revealed that a more complex mechanism, estimation of future reward,
fits the data better.
